Phylogenetics and the human microbiome.

### 2013-09-27 sbreview.tex
Commits of the morning.

Considerable attention has been given to how the microbiome influences host phenotype.
There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

@@

In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved} which uses a heuristic algorithm to optimize $F$-measure of precision and recall.
Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis.
Our group \citep{matsen2011reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.

@@

Very large data sets.

Some groups are not interested in accuracy per se.

Alignment.
\citep{eddy1998profile}
Profile hidden Markov models.

@@

@@

However, the apparent success of 16s tree based comparisons begs the question of if a better representation of the evolutionary history of the microbes would yield better predictions.
That is, if the 16S tree is incomplete and useful, what could we replace it with that would also be incomplete but even more useful?
This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict?
This representation could be simple.
For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more accurate prediction?
Would a species tree reconciled with a collection of gene trees allow for better prediction?
It is quite possible that there is nothing better given the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?

@@

Preservation of diversity.
\cite{blaser2011antibiotic,cho2012human}
"Overall these studies reveal that human microbiome data has been preserved in some coprolites, and these preserved human microbiomes match more closely to those from the rural communities than to those from cosmopolitan communities. These results suggest that the modern cosmopolitan lifestyle resulted in a dramatic change to the human gut microbiome."

@@

@@

The field has experienced a frenetic rate of expansion over the past xxx years, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
Although the human microbiome has gained a lot of attention recently, and in some cases too much attention, our microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has contributed significantly to ecology in the past, and indeed has already contributed significantly to microbial ecology; the next generation of methods may require integration of genome-level data.

>>>>

Considerable attention has been given to how the microbiome influences host phenotype.
There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.
kwashiorkor \cite{smith2013gut}.

@@

In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved} which uses a heuristic algorithm to optimize $F$-measure of precision and recall.
Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis.
Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.

@@

In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
This can be defended by saying for many applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
Furthermore, given that the underlying data is typically 16S only we can expect some tree problems even with the best methods.
Additionally, as specified below, these data sets can be very large.
There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.

Alignment is focused on creation of large curated alignments, and alignment of additional sequences to those established alignments.
There are two primary such projects: one is the SILVA project \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes \citep{desantis2006greengenes,mcdonald2011improved} database.
Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
Taking the length of 16S to be 1543 nt,
% SILVA
% 1543 / 42284
% .96353230536373096207
SILVA alignment version 115 has XXX sequences and is over 96\% gap,
% GreenGenes
% 1543 / 7682
% .79914084873730799272
while the GreenGenes 13\_5 alignment has 1,262,986 sequences and  is almost 80\% gap when aligned by SSU-Align \cite{nawrocki2009structural}
The SILVA alignment is... hand aligned?

Such large data

MSA tool specializing in 16S \cite{pruesse2012sina}.

PYNAST
ARB alignment.

\citep{eddy1998profile}
Profile hidden Markov models.

@@

Living tree project.

Given the scale of the problems, it is not surprising Bayesian phylogenetic inference is absent.

@@

However, the apparent success of 16s tree based comparisons begs the question of if a better representation of the evolutionary history of the microbes would yield better predictions.
That is, if the 16S tree is incomplete and useful, what could we replace it with that would also be necessarily incomplete but even more useful?
This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict features of underlying genomes?
This representation could be simple.
For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more accurate prediction?
Would a species tree reconciled with a collection of gene trees allow for better prediction?
It is quite possible that there is nothing better given the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?

@@

Diversity preservation is also of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity.
Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer and has made passionate appeals to preserve microbiome diversity \cite{blaser2011antibiotic,cho2012human}.
Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
\cite{tito2012insights} also cite knight group paper and global microbiome.
Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.

@@

% Companies? http://www.ribocon.com/ second genome, etc

@@

Although I have described some areas for future research for phylogeneticists, researchers should be warned that their work may not be widely accepted by the microbial ecology community.
That is, methodological changes that would be viewed as improvements in, substitution modeling for example, may not readily be taken up by the microbial ecology community.
My sense is that this is in part because the bulk of microbial ecology researchers are doctors, immunologists, and oceanographers that have an applied focus and for which computational and modeling issues are a nuisance rather than an interesting aspect.
For this reason, computational analysis is primarily done in advance or as part of established pipelines and is hardened into standard operating procedures \citep{peplies2008standard}.

Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
Although the human microbiome has gained a lot of attention recently, and in some cases too much attention, our microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has contributed significantly to ecology in the past, and indeed has already contributed significantly to microbial ecology; the next generation of methods may require integration of genome-level data.
